---
title: "Long-read_analysis"
author: "Morgane Fierville & Ali Hamraoui"
date: "2024-09-06"
output:
  html_document:
    code_folding: show
    code_download: true
    toc: true
    toc_float: true
    number_sections: false
---

<style>
body {
text-align: justify}
</style>

<!-- Set default parameters for all chunks -->
```{r, setup, include = FALSE}
set.seed(1337L)
knitr::opts_chunk$set(echo = TRUE,

                      message = FALSE,
                      warning = FALSE,
                      fold_output = FALSE, 
                      fold_plot = FALSE,
                      
                      fig.align = 'center',
                      fig.width = 20,
                      fig.height = 15)
```

```{r}
library(Seurat)
library(dplyr)
options(Seurat.object.assay.version = "v4")
```

```{r}
cell_markers <-  list(Cajal_Retzius = c('Lhx5', 'Reln', 'Snhg11', 'Meg3'),
                      GABAergic = c('Arx', 'Maf', 'Dlx2'),
                      imature_GABAergic = c('Gad2', 'Meis2', 'Gad2', 'Dlx6os1'),
                      Glutamatergic = c('Grin2b', 'Opcml', 'Camk2b', 'Mef2c'),
                      imature_Glutamatergic = c('Sox11', 'Pou3f3', 'Pou3f2',  'Neurod6', 'Neurod2'),
                      intermediate_progenitors = c('Snhg11', 'Neurog2', 'Eomes', 'Neurod1', 'Rbfox3', 'Tcf4', 'Meis2', 'H3f3b', 'Neurod6', 'Neurod2', 'Sox11', 'Pax6'),
                      cyclin_radial_glia = c('Top2a', 'Ccnd2', 'Cenpf', 'Mki67'),
                      radial_glia = c('Fabp7', 'Vim', 'Dbi' , 'Pax6'))
```

```{r}
source('../../script_audrey/cell_annot_custom.R')

rename_iso <- function(x){
        x$iso_id <-apply(x[, c("geneId","transcriptId")], 1 , paste , collapse = ".." )
        rownames(x) <- x$iso_id
        x <- x[, !colnames(x)%in%c("geneId","transcriptId","iso_id")]
        }

annotate <- function (ScObject, cell_markers) {
        if ("cell_type" %in% colnames(ScObject@meta.data)) {
          ScObject$cell_type = NULL
        }
        ScObject = cell_annot_custom(ScObject,
                                     assay = "RNA",         
                                     newname = "cell_type",  
                                     markers = cell_markers) 
        summary(ScObject$cell_type)
        return(ScObject)
}
```


```{r}
data_dir <- "./data/"
out_dir <- "./output/"
```

In this section, we load the count matrices.

### Genes

We load the genes count matrix :

load("data/illumina_data.Rdata")
gene.mtx = Illumina$scNaUmi_seq

```{r load_genes_cm}
c190_gene <- Read10X(data.dir = paste0(data_dir, "GSM3748086_190c/"))
c951_gene <- Read10X(data.dir = paste0(data_dir, "GSM3748088_951c/"))
```

### Isoforms

```{r}
c190_iso <- read.csv(paste0(data_dir,'GSM3748087_190c.isoforms.matrix.txt'), sep = '\t')
c951_iso <- read.csv(paste0(data_dir,'GSM3748089_951c.isoforms.matrix.txt'), sep = '\t')

c190_iso <- rename_iso(c190_iso)
c951_iso <- rename_iso(c951_iso)
```

### Homogeneization

We homogenize both 190c matrix annotations :
```{r homogenize}
colnames(c190_gene) = gsub("-1$", "", colnames(c190_gene))
com.cells = intersect(colnames(c190_iso), colnames(c190_gene))

c190_gene = c190_gene[, colnames(c190_gene)%in%com.cells]
c190_iso = c190_iso[, colnames(c190_iso)%in%com.cells]
```

Homogenize both 951c matrix annotations?
```{r homogenize 951c}
colnames(c951_gene) = gsub("-1$", "", colnames(c951_gene))
com.cells = intersect(colnames(c951_iso), colnames(c951_gene))

c951_gene = c951_gene[, colnames(c951_gene)%in%com.cells]
c951_iso = c951_iso[, colnames(c951_iso)%in%com.cells]
```

## Make Seurat object

We create assays for each matrix :

```{r create_assays}
c190_gene_assay = CreateAssayObject(counts = c190_gene)
c190_isoform_assay = CreateAssayObject(counts = c190_iso)
```

```{r create_assays 951c}
c951_gene_assay = CreateAssayObject(counts = c951_gene)
c951_isoform_assay = CreateAssayObject(counts = c951_iso)
```

and initialize a Seurat object :

```{r init_sobj}
assay = "RNA"
sample_name = "190c"
c190_Sobj = Seurat::CreateSeuratObject(counts = c190_gene_assay,
                                        assay = assay,
                                        project = sample_name)

c190_Sobj[["ISO"]] = c190_isoform_assay
c190_Sobj[[paste0('log_nCount_', assay)]] = log(c190_Sobj[[paste0('nCount_', assay)]])

c190_Sobj
head(rownames(c190_Sobj@assays$RNA$counts))
head(rownames(c190_Sobj@assays$ISO$counts))
```

```{r init_sobj_951c}
assay = "RNA"
sample_name = "951c"
c951_Sobj = Seurat::CreateSeuratObject(counts = c951_gene_assay,
                                        assay = assay,
                                        project = sample_name)

c951_Sobj[["ISO"]] = c951_isoform_assay
c951_Sobj[[paste0('log_nCount_', assay)]] = log(c951_Sobj[[paste0('nCount_', assay)]])

c951_Sobj
head(rownames(c951_Sobj@assays$RNA$counts))
head(rownames(c951_Sobj@assays$ISO$counts))
```

#########
## Integration of the two objects
#################################

```{r}
# put dataset into a list of two seurat objects (190c and 951c)
Sobj.list <- list(c951_Sobj, c190_Sobj)

# normalize and identify variable features for each dataset independently
Sobj.list <- lapply(X = Sobj.list, FUN = function(x) {
    x = Seurat::NormalizeData(x,
                             normalization.method = "LogNormalize")

    x = Seurat::FindVariableFeatures(x,
                                    normalization.method = "vst",
                                    nfeatures = 2000)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = Sobj.list)
```

```{r}
anchors <- FindIntegrationAnchors(object.list = Sobj.list, anchor.features = features)
seurat_obj <- IntegrateData(anchorset = anchors, new.assay.name = "integrated")
seurat_obj
```

We generate a projection based on the genes count matrix, before cells filtering.

```{r set_assay_before}
Seurat::DefaultAssay(seurat_obj) = "integrated"
```

## Projection

We generate a UMAP to visualize cells before filtering. First, we perform a PCA :

```{r pca_before, fig.width = 12, fig.height = 4}
seurat_obj = Seurat::ScaleData(seurat_obj,
                               features = rownames(seurat_obj), verbose = F)
var_features = Seurat::VariableFeatures(object = seurat_obj)

seurat_obj = Seurat::RunPCA(seurat_obj, features = var_features,
                    verbose = F, reduction.name = "RNA_pca", max_dims = 50)

Seurat::ElbowPlot(seurat_obj, ndims=100, reduction = "RNA_pca")
```

We generate a UMAP and tSNE with 10 principal components :

```{r umap_before}
ndims = 10
seurat_obj = Seurat::RunUMAP(seurat_obj,
                             reduction = "RNA_pca",
                             dims = 1:ndims,
                             seed.use = 1337L,
                             reduction.name = paste0("RNA_pca_", ndims, "_umap"))

seurat_obj = Seurat::RunTSNE(seurat_obj,
                             reduction = "RNA_pca",
                             dims = 1:ndims,
                             seed.use = 1337L,
                             reduction.name = paste0("RNA_pca_", ndims, "_tsne"))

seurat_obj
```

```{r fig.height=5, fig.width=8}
reduction = paste0("RNA_pca_", ndims, "_tsne")
DimPlot(seurat_obj, reduction = reduction)
```

## Clustering

We generate a clustering :
```{r}
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:ndims, verbose = F, reduction = 'RNA_pca')
seurat_obj <- FindClusters(seurat_obj, resolution = 0.6, verbose = F)
```

We visualize the clustering :
```{r fig.height=5, fig.width=8}
DimPlot(seurat_obj, reduction = reduction)
```

## Cell type annotation

We annotate cells for cell type using `Seurat::AddModuleScore` function.

```{r fig.height=5, fig.width=8}
seurat_obj <- annotate(seurat_obj, cell_markers)
DimPlot(seurat_obj, reduction = reduction, group.by = 'cell_type')
```

We annotate clusters :
```{r}
cell_data <- data.frame(
  cluster = Idents(seurat_obj),
  cell_type = seurat_obj$cell_type
)

major_cell_type_per_cluster <- cell_data %>%
  group_by(cluster) %>%
  summarise(cluster_cell_type = names(which.max(table(cell_type))))

seurat_obj$cluster_cell_type <- major_cell_type_per_cluster$cluster_cell_type[match(Idents(seurat_obj), major_cell_type_per_cluster$cluster)]
```

We visualize cluster annotation on the projection :

```{r fig.height=5, fig.width=8}
DimPlot(seurat_obj, reduction = "RNA_pca_10_tsne", group.by = 'cluster_cell_type', label = T)
```

## Isoforms 

Here, we are looking for events where two isoforms of the same gene are considered markers of different clusters.

We process the isoforms count matrix :

```{r process_iso}
seurat_obj = Seurat::NormalizeData(seurat_obj, assay = "ISO",
                                   normalization.method = "LogNormalize", verbose = F) 
seurat_obj = Seurat::ScaleData(seurat_obj, assay = "ISO", verbose = FALSE)
```

We generate a "multi" assay using `isoswitch` package :

```{r isoswitch_multi}
seurat_obj = isoswitch::iso_preprocess(seurat_obj, assay = "ISO", new_assay = "multi", filter_threshold = 5)
head(rownames(seurat_obj@assays$multi@counts))
```

We finally have three assays with distinct dimensions :

```{r dim_assays}
dim(seurat_obj@assays$RNA@counts)
dim(seurat_obj@assays$ISO@counts)
dim(seurat_obj@assays$multi@counts)
```

we search for gene markers. then returns genes with >1 transcripts in different clusters. (ISO_SWITCH_ALL() method uses Seurat’s FindMarkers)

```{r iso_compute_stats}
#seurat_obj@active.ident <- seurat_obj@meta.data$cluster_cell_type
Idents(seurat_obj) <- 'cluster_cell_type'
stats = isoswitch::iso_compute_stats(seurat_obj@assays$multi@counts) %>%
  dplyr::arrange(gene_id)

head(stats, n=4)
```

We plot a summary with number of genes, number of transcripts, distribution of isoforms and number of genes per cluster.

```{r plot_assay_stats, fig.height=6, fig.width=12}
isoswitch::plot_assay_stats(seurat_obj, "ISO")
```

## Save

We save the annotated and filtered Seurat object :

```{r save_seurat_obj_filtered_annotated}
saveRDS(seurat_obj, file = paste0(out_dir, "/datasets/", "seurat_obj_annotated.rds"))
```

## R session

```{r sessioninfo, echo = FALSE, fold_output = TRUE}
sessionInfo()
```