---
title: "Long-read_analysis"
author: "Morgane Fierville & Ali Hamraoui"
date: "2024-09-06"
output:
  html_document:
    code_folding: show
    code_download: true
    toc: true
    toc_float: true
    number_sections: false
---

<style>
body {
text-align: justify}
</style>

<!-- Set default parameters for all chunks -->
```{r, setup, include = FALSE}
set.seed(1337L)
knitr::opts_chunk$set(echo = TRUE,

                      message = FALSE,
                      warning = FALSE,
                      fold_output = FALSE, 
                      fold_plot = FALSE,
                      
                      fig.align = 'center',
                      fig.width = 20,
                      fig.height = 15)
```

This file is used to identify differentially expressed isoforms.

```{r library}
library(dplyr)
library(patchwork)
library(ggplot2)

.libPaths()
```

# Preparation

In this section, we set the global settings of the analysis. We will store data there :

```{r out_dir}
out_dir = "./output/"
```

We load the Seurat object :

```{r load_sobj}
seurat_obj = readRDS(paste0(out_dir, "/datasets/seurat_obj_annotated.rds"))
```

This is the projection of interest :

```{r name2D}
name2D =  "RNA_pca_10_tsne"
```

# Dataset

We visualize cells :

```{r see_cells, fig.height=5, fig.width=12, message=FALSE, warning=FALSE, paged.print=FALSE}
cell_type = Seurat::DimPlot(seurat_obj, reduction = name2D, group.by = "cluster_cell_type") +
  Seurat::NoAxes() + #Seurat::NoLegend() +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 plot.subtitle = element_text(hjust = 0.5))

clusters = Seurat::DimPlot(seurat_obj, group.by = "seurat_clusters", label = TRUE,
                           reduction = name2D) +
  Seurat::NoAxes() + ggplot2::ggtitle("Clustering") +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5))

clusters | cell_type
```

# Isoforms DE

In this section, we use `isoswitch` package to perform differential expression for isoforms.

## Between pair of clusters

We perform differential expression for isoforms, between each pair of clusters :

```{r ide}
clusters = levels(seurat_obj@active.ident)
switch_markers = isoswitch::ISO_SWITCH_ALL(seurat_obj, clusters, assay = "ISO", 
                                           min.pct = 0, logfc.threshold = 0.40)

head(switch_markers)
```

We visualize the number of DE isoforms between each cluster :

```{r plot_marker_matrix, fig.height=6, fig.width=12}
pl1 = isoswitch::plot_marker_matrix(seurat_obj, switch_markers) 
#switch_markers2 = switch_markers[!is.infinite(switch_markers$avg_log2FC),]
pl2 = isoswitch::plot_marker_score(seurat_obj, switch_markers)
pl1 | pl2
```
## Save

We save the list of DE isoforms :

```{r save_sobj}
saveRDS(switch_markers, file = paste0(out_dir, "./datasets/", "switch_markers_combined_all.rds"))
```

## R Session

```{r sessioninfo, echo = FALSE, fold_output = TRUE}
sessionInfo()
```
