---
title: "Long-read_analysis"
author: "Morgane Fierville & Ali Hamraoui"
date: "2024-09-06"
output: html_document
---

```{r}
library(Seurat)
options(Seurat.object.assay.version = "v5")
```


```{r}
cell_markers <- list(Cajal_Retzius = c('Lhx5', 'Reln', 'Snhg11', 'Meg3'),
GABAergic = c('Arx', 'Maf', 'Dlx2'),
imature_GABAergic = c('Gad2', 'Meis2', 'Gad2', 'Dlx6os1'),
Glutamatergic = c('Grin2b', 'Opcml', 'Camk2b', 'Mef2c'),
imature_Glutamatergic = c('Sox11', 'Pou3f3', 'Pou3f2',  'Neurod6', 'Neurod2'),
intermediate_progenitors = c('Snhg11', 'Neurog2', 'Eomes', 'Neurod1', 'Rbfox3', 'Tcf4', 'Meis2', 'H3f3b', 'Neurod6', 'Neurod2', 'Sox11', 'Pax6'),
cyclin_radial_glia = c('Top2a', 'Ccnd2', 'Cenpf', 'Mki67'),
radial_glia = c('Fabp7', 'Vim', 'Dbi' , 'Pax6'))

source('../../script_audrey/cell_annot_custom.R')

annotate <- function (ScObject, cell_markers) {
  if ("cell_type" %in% colnames(ScObject@meta.data)) {
    ScObject$cell_type = NULL
  }
  ScObject = cell_annot_custom(ScObject,
                               assay = "RNA",          # which assay to use to perform annotation ?
                               newname = "cell_type",  # new column name in meta.data
                               markers = cell_markers) # markers for cell type
  summary(ScObject$cell_type)
  #DimPlot(ScObject, group.by = "cell_type", cols = color_markers)
  return(ScObject)
}
```


```{r}
data_dir <- "../GSE130708_RAW/"
```

In this section, we load the count matrices.

### Genes

We load the genes count matrix :

load("data/illumina_data.Rdata")
gene.mtx = Illumina$scNaUmi_seq

```{r load_genes_cm}
c190_gene <- Read10X(data.dir = paste0(data_dir, "GSM3748086_190c/"))
c951_gene <- Read10X(data.dir = paste0(data_dir, "GSM3748088_951c/"))
```

### Isoforms

```{r}
c190_iso <- read.csv('GSM3748087_190c.isoforms.matrix.txt', sep = '\t', row.names = 2)[,-1]
c951_iso <- read.csv('GSM3748089_951c.isoforms.matrix.txt', sep = '\t', row.names = 2)[,-1]
```

### Homogeneization

We homogenize both 190c matrix annotations :
```{r homogenize}
colnames(c190_gene) = gsub("-1$", "", colnames(c190_gene))
com.cells = intersect(colnames(c190_iso), colnames(c190_gene))

c190_gene = c190_gene[, colnames(c190_gene)%in%com.cells]
c190_iso = c190_iso[, colnames(c190_iso)%in%com.cells]
```

Homogenize both 951c matrix annotations?
```{r homogenize 951c}
colnames(c951_gene) = gsub("-1$", "", colnames(c951_gene))
com.cells = intersect(colnames(c951_iso), colnames(c951_gene))

c951_gene = c951_gene[, colnames(c951_gene)%in%com.cells]
c951_iso = c951_iso[, colnames(c951_iso)%in%com.cells]
```

## Make Seurat object

We create assays for each matrix :

```{r create_assays}
c190_gene_assay = CreateAssay5Object(counts = c190_gene)
c190_isoform_assay = CreateAssay5Object(counts = c190_iso)
```

```{r create_assays 951c}
c951_gene_assay = CreateAssay5Object(counts = c951_gene)
c951_isoform_assay = CreateAssay5Object(counts = c951_iso)
```

and initialize a Seurat object :

```{r init_sobj}
assay = "RNA"
sample_name = "190c"
c190_Sobj = Seurat::CreateSeuratObject(counts = c190_gene_assay,
                                        assay = assay,
                                        project = sample_name)

c190_Sobj[["ISO"]] = c190_isoform_assay
c190_Sobj[[paste0('log_nCount_', assay)]] = log(c190_Sobj[[paste0('nCount_', assay)]])

c190_Sobj
head(rownames(c190_Sobj@assays$RNA$counts))
head(rownames(c190_Sobj@assays$ISO$counts))
```

```{r init_sobj_951c}
assay = "RNA"
sample_name = "951c"
c951_Sobj = Seurat::CreateSeuratObject(counts = c951_gene_assay,
                                        assay = assay,
                                        project = sample_name)

c951_Sobj[["ISO"]] = c951_isoform_assay
c951_Sobj[[paste0('log_nCount_', assay)]] = log(c951_Sobj[[paste0('nCount_', assay)]])

c951_Sobj
head(rownames(c951_Sobj@assays$RNA$counts))
head(rownames(c951_Sobj@assays$ISO$counts))
```


```{r}
seurat_obj <- merge(c951_Sobj, c190_Sobj)
seurat_obj
```

# Before filtering

We generate a projection based on the genes count matrix, before cells filtering.

```{r set_assay_before}
Seurat::DefaultAssay(seurat_obj) = "RNA"
```

## Normalization

We normalize the genes count matrix.

```{r normalization}
seurat_obj = Seurat::NormalizeData(seurat_obj,
                                   normalization.method = "LogNormalize",
                                   assay = "RNA")

seurat_obj = Seurat::FindVariableFeatures(seurat_obj,
                                          assay = "RNA",
                                          normalization.method = "vst",
                                          nfeatures = 3000)
seurat_obj
```

## Projection

We generate a UMAP to visualize cells before filtering. First, we perform a PCA :

```{r pca_before, fig.width = 12, fig.height = 4}
seurat_obj = Seurat::ScaleData(seurat_obj,
                               features = rownames(seurat_obj), verbose = F)
var_features = Seurat::VariableFeatures(object = seurat_obj)

seurat_obj = Seurat::RunPCA(seurat_obj, features = var_features,
                    verbose = F, reduction.name = "RNA_pca", max_dims = 100)

Seurat::ElbowPlot(seurat_obj, ndims=100, reduction = "RNA_pca")
```

We generate a UMAP with 10 principal components :

```{r umap_before}
ndims = 10
seurat_obj = Seurat::RunUMAP(seurat_obj,
                             reduction = "RNA_pca",
                             dims = 1:ndims,
                             seed.use = 1337L,
                             reduction.name = paste0("RNA_pca_", ndims, "_umap"))

seurat_obj = Seurat::RunTSNE(seurat_obj,
                             reduction = "RNA_pca",
                             dims = 1:ndims,
                             seed.use = 1337L,
                             reduction.name = paste0("RNA_pca_", ndims, "_tsne"))

seurat_obj
```

```{r}
DimPlot(seurat_obj, reduction = "RNA_pca_10_tsne")
```


```{r}
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = F, reduction = 'RNA_pca')
seurat_obj <- FindClusters(seurat_obj, resolution = 0.5, verbose = F)
```


```{r}
DimPlot(seurat_obj, reduction = "RNA_pca_10_tsne")
```

```{r}
seurat_obj <- JoinLayers(seurat_obj)
seurat_obj <- annotate(seurat_obj, cell_markers)
```

```{r}
DimPlot(seurat_obj, reduction = "RNA_pca_10_tsne", group.by = 'cell_type')
```

```{r}
cell_data <- data.frame(
  cluster = Idents(seurat_obj),
  cell_type = seurat_obj$cell_type
)

major_cell_type_per_cluster <- cell_data %>%
  group_by(cluster) %>%
  summarise(major_cell_type = names(which.max(table(cell_type))))

seurat_obj$major_cell_type <- major_cell_type_per_cluster$major_cell_type[match(Idents(seurat_obj), major_cell_type_per_cluster$cluster)]
```

```{r}
DimPlot(seurat_obj, reduction = "RNA_pca_10_tsne", group.by = 'major_cell_type', label = T)
```
